Using Fundamental Measure Theory to Treat the Correlation Function of the Inhomogeneous Hard-Sphere Fluid

### 2012-08-14 papers/contact/paper.tex
content edit of beginning of contact paper Tested everything on node06.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We investigate the value of the correlation function an
  inhomogeneous hard-sphere fluid at contact $g_\sigma$.  This
  quantity plays a critical role in Statistical Associating Fluid
  Theory (SAFT), which is the basis of a number of recent classical
  density functionals.  We define two averaged values for the
  correlation function at contact, and derive formulas for each of
  them White Bear version of the Fundamental Measure Theory
  functional~\cite{roth2002whitebear} based on an assumption of
  thermodynamic consistency. We have tested these functionals and two
  existing functionals~\cite{yu2002fmt-dft-inhomogeneous-associating,
    gross2009density} against Monte Carlo simulations.  We find that
  excellent agreement with the Monte Carlo results for our
  asymmetrically averaged correlation function, and less impressive
  results for the two existing functionals and our symmetrically
  averaged correlation function.
\end{abstract}

@@

There has been considerable recent interest in using Statistical
Associating Fluid Theory (SAFT) to construct classical density
functionals to describe associating
fluids\cite{\saftlocaldft,\saftnonlocaldft}.  This approach has been
successful in qualitatively describing the dependence of surface
tension on temperature.  Unfortunately, most of these constructed
functionals\cite{\saftlocaldft} cannot be used to study arbitrary
inhomogeneous density distributions, because they use local density
approximations that are only valid for densities that vary slowly over
molecular length scales.  A key input to those SAFT-based functionals
that can handle sharply varying densities is the correlation function
evaluated at contact.

Constructing a classical density functional based on SAFT involves
rewriting each term in the free energy as a functional of an
inhomogeneous density.  The SAFT free energy consists of four terms:
\begin{align}
  A_\textit{SAFT} &= A_\textit{ideal} + A_\textit{HS} + A_\textit{chain} + A_\textit{disp} + A_\textit{assoc}
\end{align}
where $A_\textit{ideal}$ is the ideal-gas free energy, $A_\textit{HS}$
is the excess free energy of a hard-sphere fluid of monomers,
$A_\textit{chain}$ is the contribution to the free energy from the
chaining effect in a polymeric fluid, $A_\textit{disp}$ is the
dispersion contribution to the free energy, and $A_\textit{assoc}$ is
the free energy contribution from \emph{association}, which is to say,
hydrogen bonding.  The chain and association contributions to the SAFT
free energy are of particular interest for this paper, since each uses
as input the contact value of the correlation function.
%
Yu and Wu introduced in 2002 a functional for the association term of
the free energy, which included a functional for the contact value of
the correlation function (see
Section~\ref{sec:yuwu})\cite{yu2002fmt-dft-inhomogeneous-associating},
which has subsequently been used in other
functionals\cite{fu2005vapor-liquid-dft, bryk2006density}.
%% Each of these terms has some dependence on the
%% inhomogeneity in the density.  The ideal gas is purely local, and in
%% fact must be the only purely local term in the free energy in order
%% for the functional to satisfy the contact-value theorem.  The
%% hard-sphere contribution contribution to the free energy has been
%% thoroughly studied, and is well approximated by the White Bear version
%% of the Fundamental-Measure Theory (FMT)
%% functional\cite{roth2002whitebear}.
Two functionals for the chain contribution have recently been
introduced\cite{bryk2006density, gross2009density}, one which uses the
contact value of the correlation function of Yu and
Wu\cite{bryk2006density}, while the other introduces a new functional
for the contact value of the correlation
function (see Section~\ref{sec:gross})\cite{gross2009density}.
%% The dispersion term is long-range and thus has significant
%% dependence on the density distribution, but because its relatively
%% weak position dependence, is amenable to mean-field approximations.
Given these different approaches, it seems worthwhile to examine this
property of the hard-sphere fluid through direct simulation, in order
to establish the advantages and disadvantages of each approach.

We first discuss two possibilities for the definition of contact
density.  The first involves a more symmetric interpretation and so we
call it the '$S$ case'.  The second involves a sphere-centered,
assymetrical interpretation and so we call it the '$A$ case'.  We
present where in the SAFT free energy formulation these contact
densities will contribute.  We next present a thermodynamic method
that we will use to calculate the above contact densities.  This
method involves the use of surface pressure and the minimization of
the excess free energy refered to above.  We use the White Bear and
White Bear Mark II formulations of these free energy terms and take
neccesary derivitives in the concpetual spirit of the seperately
defined $S$ an $A$ case contact densities.  Finally we compare the
resulting contact densities with those of Gross, those of Yu and Wu,
and as well with our own Monte Carlo simulations.

@@

We will define our terms by making use of the two-particle density
$n^{(2)}(\rr_1,\rr_2)$, which gives the probability per unit volume of
finding one particle at position $\rr_1$ and the other at position
$\rr_2$.  The pair correlation function is defined by
\begin{align}
  g(\rr_1,\rr_2) &\equiv \frac{n^{(2)}(\rr_1,\rr_2)}{n(\rr_1)n(\rr_2)}
\end{align}
In a homogeneous fluid, the pair correlation only depends on the
distance $|\rr_1-\rr_2|$, and can be expressed as a function of a
single variable, with the contact value being when that distance is
the diameter $\sigma$.  It is desirable for reasons of efficiency to limit CDFT
functionals to one-center convolutions, which leads us to seek a
simplified expression for the contact value of the correlation
function---which is the same as the contact value of the cavity
correlation function for hard spheres.
In a system with an inhomogeneous density, we seek a \emph{local}
value for $g_\sigma$.  There are two reasonable options for defining
such a local function: a symmetric formulation (which we refer to as $S$) and an
asymmetric formulation (which we refer to as $A$).

For the symmetric $S$ case:
\begin{align}
  g^S_\sigma(\rr) &= \frac{1}{n_0(\rr)^2}\int n^{(2)}(\rr - \rr', \rr
  + \rr')
  \frac{\delta(|\rr'| -\sigma/2)}{\pi\sigma^2}d\rr' \label{eq:gS} \\
  n_0(\rr) &= \int n(\rr')\frac{\delta(|\rr-\rr'|-\sigma/2)}{\pi\sigma^2} d\rr'
\end{align}
This functional gives a value averaged over spheres that touch at the
position $\rr$.  The density $n_0$ is one of the fundamental measures
of Fundamental Measure Theory (FMT), and is ideal for treating
touching spheres, as illustrated in Figure~\ref{fig:n0}.

@@

And for the asymmetric $A$ case:
\begin{align}
  g^A_\sigma(\rr) &= \frac{1}{n(\rr)n_A(\rr)}
  \int n^{(2)}(\rr, \rr + \rr')
  \frac{\delta(|\rr'| - \sigma)}{4\pi\sigma^2}d\rr' \label{eq:gA} \\
  n_A(\rr) &= \int n(\rr')
  \frac{\delta(|\rr-\rr'|-\sigma)}{4\pi\sigma^2} d\rr' \label{eq:nA}
\end{align}
In this case, the two-particle density is averaged over spheres
touching a sphere that is at the given position~$\rr$.

In the process of defining these two averaged correlation functions,
we also defined two averaged densities, $n_0(\rr)$ and $n_A(\rr)$,
which are the density of spheres available to be in
contact.  In the homogeneous limit, $n_0 = n_A = n$.  It is an
open question, which of these averages will be more useful in any
particular functional.  Suffice to say, either average is a
\emph{possible} way to convert a function that is defined for a
homogeneous system to a functional that is applicable to inhomogeneous
systems.

Yu and Wu introduce an approximation for
$g^S$\cite{yu2002fmt-dft-inhomogeneous-associating}, which we will
discus in Section~\ref{sec:yuwu}.  Gross introduced an
approximation for $g^A$\cite{gross2009density}, which we will
introduce in Section~\ref{sec:gross}.  We will also introduce our own
approximations for both $g^A$ and $g^S$ which are based on the
assumption of thermodynamic consistency within FMT, in
Sections~\ref{sec:g-S} and~\ref{sec:g-A}.  We will report on the accuracy of
these approximations by comparing with Monte Carlo simulations of the
inhomogeneous hard-sphere fluid.  We will focus on the contact
densities, rather than the correlation function itself, because these
are simpler, and because they are the quantity that you want to
``average''.  i.e. if you were to average $g$ itself, you'd want to
weight that average based on the density, which would mean averaging
$n^{(2)}$.

\subsection{Association free energy of SAFT}

\fixme{Integrate the following equations into this section...}
\begin{align}
  \frac{A_\textit{chain}}{kT} &= -(m-1) n \left(\ln\left(n g_\sigma \right)-1\right)
\end{align}
\begin{align}
  \frac{A_\textit{assoc}}{kT} &= \sum_i n \left(\ln X_i - \frac12 X_i + \frac12\right) \\
  X_i &= \frac{1}{1 + \sum_j n g_\sigma X_j\kappa_{ij} \left(e^{\beta \epsilon_{ij}}-1\right)}
\end{align}

\newcommand\epsilonassoc{\ensuremath{\varepsilon_\textit{AB}}}
\newcommand\kappaassoc{\ensuremath{\kappa_\textit{AB}}}
\newcommand\ncontact{\ensuremath{n_\textit{contact}}}

The free energy term in Statistical Associating Fluid Teory (SAFT)
from which it derives its name is the association term, which accounts
for hydrogen bonding.  Hydrogen bonds are modeled as attractive
patches (``association sites'') on the surface of hard spheres.  These
sites represent protons or electron lone pairs, and have an attractive
energy $\epsilonassoc$ when two molecules are oriented such that the
proton of one overlaps with the lone pair of the other.  The volume
over which this interaction occurs is $\kappaassoc$, giving the
association term in the free energy has two empirical parameters fit
to experimental data.

The association free energy per unit volume has the form
\begin{align}
  f_\text{assoc} &= k_BT n\sum_A 
                  \left(\ln X^A - \frac{X^A}{2} + \frac12\right)
\end{align}
where the summation is over the association sites, and $X^A$ is the
fraction of association sites \emph{not} hydrogen-bonded.  The
fractions $X^A$ are determined by the self-consistent equations
\begin{align}
  X^A &= \frac{1}
  {1 + \sum_B n X^B \Delta^{AB}}\label{eq:X}
  \\
  \Delta^{AB} &= g(\sigma) \kappaassoc\left( e^{\epsilonassoc/kT} -
  1 \right)\label{eq:delta}
\end{align}
where $g(\sigma)$ is the correlation function evaluated at contact.
The product of density with correlation function evaluated at contact
gives the contact density, when we combine Equations~\ref{eq:X} and
\ref{eq:delta}:
\begin{align}
  X^A &= \frac{1}
  {1 + \sum_B \ncontact\kappaassoc X^B\left( e^{\epsilonassoc/kT} -
  1 \right)}
\end{align}
where the product $\ncontact\kappaassoc$ is the expected number of
molecules present in the association volume in the absence of the
association interaction.  In the SAFT-VR
model\cite{gil-villegas-1997-SAFT-VR}, this contact density is found
by adding a perturbative correction to the contact value for a hard
sphere fluid.

@@

In order to motivate the derivation of our functional for contact
density, we begin by deriving the well-known contact density for a
homogeneous hard-sphere fluid from the Carnahan-Starling free energy.
The contact density is most easily computed from the contact density
theorem, which states that the pressure on any hard surface is
determined by the density at contact:
\begin{align}
  p &= k_BT n_\textit{contact}.
\end{align}
Since we are interested in the contact density at the surface of the
hard spheres, all we need compute is the pressure on that surface, and
we'll have our answer.  The pressure on a hard sphere can be readily
computed from the dependence of the Carnahan-Starling free energy on
hard sphere radius.
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
where $\eta \equiv \frac{4\pi}{24} \sigma^3 n$ is the filling fraction.  We
can thus readily compute the derivative of free energy with respect to
hard-sphere radius:
\begin{align}
  \frac{dA_{HS}}{dR} &= 2\frac{dA_{HS}}{d\eta} \frac{d\eta}{d\sigma} \\
  \derivation{
    &= Nk_BT \left( \frac{4 - 6\eta}{(1-\eta)^2} + 2 \frac{4\eta - 3\eta^2}{(1-\eta)^3} \right) \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 4\eta - 6\eta + 6\eta^2 + 8\eta - 6\eta^2}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
  }
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{6 \eta}{\sigma} \label{eq:dAhsdR}
\end{align}
This derivative gives us the force on \emph{all} the hard
spheres---since we're changing all their radii at once.  To compute
the pressure on the spheres, we just need to divide by the total area,
which means dividing by $N 4\pi \sigma^2$, since the relevant area
is the area over which the molecules can make contact.
\begin{align}
  p_{HS} &= \frac{1}{N 2\pi \sigma^2} \frac{dA_{HS}}{d\sigma} \\
  \derivation{
    &= \frac14 \frac{1}{N 4\pi R^2} Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \\
    &= \frac{3}{4\pi R^3} \eta k_BT \frac{1 -
      \frac{\eta}2}{(1-\eta)^3} \\
  }
  &= k_BT n \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Using the contact-value theorem, we can thus find the contact density
and the well-known correlation function evaluated at contact.
\begin{align}
  n_\textit{contact} &= n \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  g_{HS}(\sigma) &= \frac{1 - \frac{\eta}2}{(1-\eta)^3} \label{eq:cs-g}
\end{align}
Thus we can readily derive the contact density for the homogeneous
hard-sphere fluid.  Extending this to the inhomogeneous fluid will
require that we use density-functional theory.

@@

When we consider an inhomogeneous system, we may prefer to ask for
\emph{which} sphere we are interested in computing the contact
density.  One way to answer this is to specify a sphere at a fixed
position $\mathbf{r}$.  By making the hard-sphere radius dependent on
position, we can compute the pressure on the hard spheres located at
$\mathbf{r}$.  Since the hard-sphere radius enters the FMT free energy
by means of the fundamental measures, this requires that we adjust the
convolutions to have a radius $R$ which depends on the location of the
sphere:
\begin{align}
  n_3^A(\rr) &= \int n(\rr')\Theta(|\rr-\rr'|-\sigma(\rr')/2)d\rr' \label{eq:n3A}
\end{align}
Once we have done this---and similarly for the other fundamental
measures---we derive a formula for the correlation function
$g_\sigma^A(\rr)$ by finding the pressure on the surface of any hard
spheres located at position $\rr$, and using the contact-density
theorem to extract the contact density, from which we can establish
the averaged correlation function at contact.
\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n(\rr) 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:p_{HS}^A}\\
  \ncontact(\rr) &= \frac{1}{n(\rr) k_BT 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \\
  g_\sigma^A(\rr) &= \frac{\ncontact(\rr)}{n_A(\rr)} \\
  &= \frac{1}{n(\rr) n_A(\rr)}\frac{1}{ k_BT 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
The functional derivative $2 \frac{\delta A_{HS}}{\delta \sigma(\mathbf{r})}$
in the above formula must be computed with the asymmetric formulation
for the fundamental measures.  In practice, we use the approximate
White Bear functional to find an approximate expression for
$g_\sigma^A$, which is given in Appendix~\ref{appendix:g-A}.  This
expression will involve convolutions of derivatives of the free
energy, making this formulation computationally more expensive than
the free energy itself.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We investigate the value of the correlation function an
  inhomogeneous hard-sphere fluid at contact $g_\sigma$.  This
  quantity plays a critical role in Statistical Associating Fluid
  Theory (SAFT), which is the basis of a number of recent classical
  density functionals.  We define two averaged values for the
  correlation function at contact, and derive formulas for each of
  them from the White Bear version of the Fundamental Measure Theory
  functional~\cite{roth2002whitebear}, based on an assumption of
  thermodynamic consistency. We have tested these functionals and two
  existing functionals~\cite{yu2002fmt-dft-inhomogeneous-associating,
    gross2009density} against Monte Carlo simulations.  We find
  excellent agreement between the Monte Carlo data and our
  asymmetrically averaged correlation function, and less impressive
  agreement with the other three correlation functionals.
\end{abstract}

@@

There has been considerable recent interest in using Statistical
Associating Fluid Theory (SAFT) to construct classical density
functionals to describe associating
fluids\cite{\saftlocaldft,\saftnonlocaldft}.  This approach has been
successful in qualitatively describing the dependence of surface
tension on temperature.
%
%% Unfortunately, most of these constructed
%% functionals\cite{\saftlocaldft} cannot be used to study arbitrary
%% inhomogeneous density distributions, because they use local density
%% approximations that are only valid for densities that vary slowly over
%% molecular length scales.
%
A key input to SAFT-based functionals is the correlation function
evaluated at contact.  Although two recent works have introduced
approximate functionals for this
quantity\cite{yu2002fmt-dft-inhomogeneous-associating,
  gross2009density}, there has yet to be a careful study of the
correlation function at contact in the inhomogeneous hard-sphere
fluid.

Constructing a classical density functional based on SAFT involves
rewriting each term in the free energy as a functional of an
inhomogeneous density.  The SAFT free energy consists of four terms:
\begin{align}
  A_\textit{SAFT} &= A_\textit{ideal} + A_\textit{HS} + A_\textit{chain} + A_\textit{disp} + A_\textit{assoc}
\end{align}
where $A_\textit{ideal}$ is the ideal-gas free energy, $A_\textit{HS}$
is the excess free energy of a hard-sphere fluid, $A_\textit{chain}$
is the free energy of formation for polymeric fluids,
$A_\textit{disp}$ is the dispersion contribution to the free energy,
and $A_\textit{assoc}$ is the free energy contribution from
\emph{association}, which is to say, hydrogen bonding.  The chain and
association contributions to the SAFT free energy are of particular
interest for this paper, since each requires as input the contact
value of the correlation function.
%
Yu and Wu introduced in 2002 a functional for the association term of
the free energy, which included a functional for the contact value of
the correlation function (see
Section~\ref{sec:yuwu})\cite{yu2002fmt-dft-inhomogeneous-associating},
which has subsequently been used in the development of other
SAFT-based functionals\cite{fu2005vapor-liquid-dft, bryk2006density}.
%% Each of these terms has some dependence on the
%% inhomogeneity in the density.  The ideal gas is purely local, and in
%% fact must be the only purely local term in the free energy in order
%% for the functional to satisfy the contact-value theorem.  The
%% hard-sphere contribution contribution to the free energy has been
%% thoroughly studied, and is well approximated by the White Bear version
%% of the Fundamental-Measure Theory (FMT)
%% functional\cite{roth2002whitebear}.
Two functionals for the chain contribution have recently been
introduced\cite{bryk2006density, gross2009density}, one which uses the
contact value of the correlation function of Yu and
Wu\cite{bryk2006density}, while the other introduces a new functional
for the contact value of the correlation
function (see Section~\ref{sec:gross})\cite{gross2009density}.
%% The dispersion term is long-range and thus has significant
%% dependence on the density distribution, but because its relatively
%% weak position dependence, is amenable to mean-field approximations.
Given these different approaches, it seems valuable to examine this
property of the hard-sphere fluid through direct simulation, in order
to establish the advantages and disadvantages of each approach.

In this paper we introduce two definitions for the locally averaged
correlation function of an inhomogeneous system.  The first is more
symmetric (which we call the $S$ case), while the second involves a
sphere-centered, assymetrical interpretation (the $A$ case).  Given
these definitions, we will present a thermodynamic approach for
evaluating these these correlation functions directly from the free
energy functional itself.  We will then explain how the correlation
function of Yu and Wu and that of Gross compare with our
definitions---the former is a variant of the $S$ correlation function,
while the latter is an $A$ function.  Finally, we will compare these
four approximations with exact Monte-Carlo simulations for the
hard-sphere fluid in at a variety of hard-wall surfaces.

@@

We will define our terms by making use of the two-particle density
$n^{(2)}(\rr_1,\rr_2)$, which gives the probability per unit volume
squared of finding one particle at position $\rr_1$ and the other at
position $\rr_2$.  The pair correlation function is defined by
\begin{align}
  g(\rr_1,\rr_2) &\equiv \frac{n^{(2)}(\rr_1,\rr_2)}{n(\rr_1)n(\rr_2)}
\end{align}
In a homogeneous fluid, the pair correlation only depends on the
distance $|\rr_1-\rr_2|$, and can be expressed as a function of a
single variable, with the contact value being when that distance is
the diameter $\sigma$.  It is desirable for reasons of efficiency to limit CDFT
functionals to one-center convolutions, which leads us to seek a
simplified expression for the contact value of the correlation
function---which is the same as the contact value of the cavity
correlation function for hard spheres.
In a system with an inhomogeneous density, we seek a \emph{local}
value for $g_\sigma$.  There are two reasonable options for defining
such a local function: a symmetric formulation (which we refer to as $S$) and an
asymmetric formulation (which we refer to as $A$).

For the symmetric $S$ case, the correlation function at contact is
given by:
\begin{align}
  g^S_\sigma(\rr) &= \frac{1}{n_0(\rr)^2}\int n^{(2)}(\rr - \rr', \rr
  + \rr')
  \frac{\delta(|\rr'| -\sigma/2)}{\pi\sigma^2}d\rr' \label{eq:gS}
\end{align}
where density $n_0$ is one of the fundamental measures of Fundamental
Measure Theory (FMT), and is ideal for treating touching spheres, as
illustrated in Figure~\ref{fig:n0}.
\begin{align}
  n_0(\rr) &= \int n(\rr')\frac{\delta(|\rr-\rr'|-\sigma/2)}{\pi\sigma^2} d\rr'
\end{align}
This functional gives a value averaged over all spheres that touch at
the position $\rr$.

@@

In contrast, the asymmetrically averaged $A$ correlation function is
given by
\begin{align}
  g^A_\sigma(\rr) &= \frac{1}{n(\rr)n_A(\rr)}
  \int n^{(2)}(\rr, \rr + \rr')
  \frac{\delta(|\rr'| - \sigma)}{4\pi\sigma^2}d\rr' \label{eq:gA}
\end{align}
where the density $n_A(\rr)$ is similar to $n_0$, but measures the
density of spheres that could be touching a sphere that is located at
point $\rr$.
\begin{align}
  n_A(\rr) &= \int n(\rr')
  \frac{\delta(|\rr-\rr'|-\sigma)}{4\pi\sigma^2} d\rr' \label{eq:nA}
\end{align}
Thus $g_\sigma^A$ corresponds to an average of the two-particle
density over spheres touching a sphere that is located at the
position~$\rr$.

%% In the process of defining these two averaged correlation functions,
%% we also defined two averaged densities, $n_0(\rr)$ and $n_A(\rr)$,
%% which are the density of spheres available to be in
%% contact.  In the homogeneous limit, $n_0 = n_A = n$.  It is an
%% open question, which of these averages will be more useful in any
%% particular functional.  Suffice to say, either average is a
%% \emph{possible} way to convert a function that is defined for a
%% homogeneous system to a functional that is applicable to inhomogeneous
%% systems.

%% Yu and Wu introduce an approximation for
%% $g^S$\cite{yu2002fmt-dft-inhomogeneous-associating}, which we will
%% discus in Section~\ref{sec:yuwu}.  Gross introduced an
%% approximation for $g^A$\cite{gross2009density}, which we will
%% introduce in Section~\ref{sec:gross}.  We will also introduce our own
%% approximations for both $g^A$ and $g^S$ which are based on the
%% assumption of thermodynamic consistency within FMT, in
%% Sections~\ref{sec:g-S} and~\ref{sec:g-A}.  We will report on the accuracy of
%% these approximations by comparing with Monte Carlo simulations of the
%% inhomogeneous hard-sphere fluid.  We will focus on the contact
%% densities, rather than the correlation function itself, because these
%% are simpler, and because they are the quantity that you want to
%% ``average''.  i.e. if you were to average $g$ itself, you'd want to
%% weight that average based on the density, which would mean averaging
%% $n^{(2)}$.

%% \subsection{Association free energy of SAFT}

%% \fixme{Integrate the following equations into this section...}
%% \begin{align}
%%   \frac{A_\textit{chain}}{kT} &= -(m-1) n \left(\ln\left(n g_\sigma \right)-1\right)
%% \end{align}
%% \begin{align}
%%   \frac{A_\textit{assoc}}{kT} &= \sum_i n \left(\ln X_i - \frac12 X_i + \frac12\right) \\
%%   X_i &= \frac{1}{1 + \sum_j n g_\sigma X_j\kappa_{ij} \left(e^{\beta \epsilon_{ij}}-1\right)}
%% \end{align}

%% \newcommand\epsilonassoc{\ensuremath{\varepsilon_\textit{AB}}}
%% \newcommand\kappaassoc{\ensuremath{\kappa_\textit{AB}}}
\newcommand\ncontact{\ensuremath{n_\textit{contact}}}

%% The free energy term in Statistical Associating Fluid Teory (SAFT)
%% from which it derives its name is the association term, which accounts
%% for hydrogen bonding.  Hydrogen bonds are modeled as attractive
%% patches (``association sites'') on the surface of hard spheres.  These
%% sites represent protons or electron lone pairs, and have an attractive
%% energy $\epsilonassoc$ when two molecules are oriented such that the
%% proton of one overlaps with the lone pair of the other.  The volume
%% over which this interaction occurs is $\kappaassoc$, giving the
%% association term in the free energy has two empirical parameters fit
%% to experimental data.

%% The association free energy per unit volume has the form
%% \begin{align}
%%   f_\text{assoc} &= k_BT n\sum_A 
%%                   \left(\ln X^A - \frac{X^A}{2} + \frac12\right)
%% \end{align}
%% where the summation is over the association sites, and $X^A$ is the
%% fraction of association sites \emph{not} hydrogen-bonded.  The
%% fractions $X^A$ are determined by the self-consistent equations
%% \begin{align}
%%   X^A &= \frac{1}
%%   {1 + \sum_B n X^B \Delta^{AB}}\label{eq:X}
%%   \\
%%   \Delta^{AB} &= g(\sigma) \kappaassoc\left( e^{\epsilonassoc/kT} -
%%   1 \right)\label{eq:delta}
%% \end{align}
%% where $g(\sigma)$ is the correlation function evaluated at contact.
%% The product of density with correlation function evaluated at contact
%% gives the contact density, when we combine Equations~\ref{eq:X} and
%% \ref{eq:delta}:
%% \begin{align}
%%   X^A &= \frac{1}
%%   {1 + \sum_B \ncontact\kappaassoc X^B\left( e^{\epsilonassoc/kT} -
%%   1 \right)}
%% \end{align}
%% where the product $\ncontact\kappaassoc$ is the expected number of
%% molecules present in the association volume in the absence of the
%% association interaction.  In the SAFT-VR
%% model\cite{gil-villegas-1997-SAFT-VR}, this contact density is found
%% by adding a perturbative correction to the contact value for a hard
%% sphere fluid.

\subsection{Fundamental-Measure Theory}

We use the White Bear version of the Fundamental-Measure Theory~(FMT)
functional published in reference~\cite{roth2002whitebear}.  The FMT
functional describes the excess free energy of a hard-sphere fluid.
This particular FMT reduces to the Carnahan-Starling equation of state
for homogeneous systems.
\begin{equation}
A_\textit{HS}[n] = k_B T \int \left(\Phi_1(\rr) + \Phi_2(\rr) + \Phi_3(\rr)\right) d\rr \; ,
\end{equation}
with integrands
\begin{align}
\Phi_1 &= -n_0 \ln\left( 1 - n_3\right)\\
\Phi_2 &= \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3} \\
\Phi_3 &= (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
  n_3 + (1-n_3)^2 \ln(1-n_3)
}{
  36\pi n_3^2\left( 1 - n_3 \right)^2
} ,
\end{align}
using the weighted densities
\begin{align}
  n_3(\rr) &= \int n(\rr') \Theta(\left|\rr - \rr'\right| - \sigma/2) d\rr' \\
  n_2(\rr) &= \int n(\rr') \delta(\left|\rr - \rr'\right| - \sigma/2) d\rr' \\
  \mathbf{n}_{2V}(\rr) &= \int n(\rr') \delta(\left|\rr - \rr'\right| - \sigma/2) \frac{\rr-\rr'}{|\rr-\rr'|}d\rr'
\end{align}
\begin{align}
  \mathbf{n}_{V1} = \frac{\mathbf{n}_{V2}}{2\pi \sigma}, \quad
  n_1 &= \frac{n_2}{2\pi \sigma} , \quad
  n_0 = \frac{n_2}{\pi \sigma^2}
\end{align}

@@

In order to motivate the derivation of our functional for the value of
the correlation function at contact, we begin by deriving the
well-known value of $g_\sigma$ for a homogeneous hard-sphere fluid
from the Carnahan-Starling free energy.  The contact value of the
correlation function density is most easily found by using the
contact-value theorem, which states that the pressure on any hard
surface is determined by the density at contact:
\begin{align}
  p &= k_BT n_\textit{contact} \\
  &= k_BT n g_\sigma
\end{align}
Since we are interested in the correlation function at the surface of
the hard spheres, all we need compute is the pressure on that surface,
and we'll have our answer.  The pressure on a hard sphere can be
readily computed from the dependence of the Carnahan-Starling free
energy on hard sphere radius.
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
where $\eta \equiv \frac{4\pi}{24} \sigma^3 n$ is the filling fraction.  We
can thus readily compute the derivative of free energy with respect to
hard-sphere radius:
\begin{align}
  \frac{dA_{HS}}{dR} &= 2\frac{dA_{HS}}{d\eta} \frac{d\eta}{d\sigma} \\
  \derivation{
    &= Nk_BT \left( \frac{4 - 6\eta}{(1-\eta)^2} + 2 \frac{4\eta - 3\eta^2}{(1-\eta)^3} \right) \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 4\eta - 6\eta + 6\eta^2 + 8\eta - 6\eta^2}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
  }
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{6 \eta}{\sigma} \label{eq:dAhsdR}
\end{align}
This derivative gives us the force on \emph{all} the hard
spheres---since we're changing all their radii at once.  To compute
the pressure on the spheres, we just need to divide by the total area,
which means dividing by $N 4\pi \sigma^2$, since the relevant area
is the area over which the molecules can make contact.
\begin{align}
  p_{HS} &= \frac{1}{N 2\pi \sigma^2} \frac{dA_{HS}}{d\sigma} \\
  \derivation{
    &= \frac14 \frac{1}{N 4\pi R^2} Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \\
    &= \frac{3}{4\pi R^3} \eta k_BT \frac{1 -
      \frac{\eta}2}{(1-\eta)^3} \\
  }
  &= k_BT n \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Using the contact-value theorem, we can thus find the contact density
and the well-known correlation function evaluated at contact.
\begin{align}
  n_\textit{contact} &= n \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  g_{HS}(\sigma) &= \frac{1 - \frac{\eta}2}{(1-\eta)^3} \label{eq:cs-g}
\end{align}
Thus we can readily derive the contact density for the homogeneous
hard-sphere fluid.  Extending this to the inhomogeneous fluid will
require that we use density-functional theory.

@@

When we consider an inhomogeneous system, we may prefer to ask for
\emph{which} sphere we are interested in computing the contact
density.  One way to answer this is to specify a sphere at a fixed
position $\mathbf{r}$.  By making the hard-sphere radius dependent on
position, we can compute the pressure on the hard spheres located at
$\mathbf{r}$.  Since the hard-sphere radius enters the FMT free energy
by means of the fundamental measures, this requires that we adjust the
convolutions to have a diameter $\sigma$ which depends on the location
of the sphere:
\begin{align}
  n_3^A(\rr) &= \int n(\rr')\Theta(|\rr-\rr'|-\sigma(\rr')/2)d\rr' \label{eq:n3A}
\end{align}
Once we have done this---and similarly for the other fundamental
measures---we derive a formula for the correlation function
$g_\sigma^A(\rr)$ by finding the pressure on the surface of any hard
spheres located at position $\rr$, and using the contact-density
theorem to extract the contact density, from which we can establish
the averaged correlation function at contact.
\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n(\rr) 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:p_{HS}^A}\\
  \ncontact(\rr) &= \frac{1}{n(\rr) k_BT 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \\
  g_\sigma^A(\rr) &= \frac{\ncontact(\rr)}{n_A(\rr)} \\
  &= \frac{1}{n(\rr) n_A(\rr)}\frac{1}{ k_BT 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
The functional derivative $2 \frac{\delta A_{HS}}{\delta \sigma(\mathbf{r})}$
in the above formula must be computed with the asymmetric formulation
for the fundamental measures.  In practice, we use the approximate
White Bear functional to find an approximate expression for
$g_\sigma^A$, which is given in Appendix~\ref{appendix:g-A}.  This
expression will involve convolutions of derivatives of the free
energy, making this formulation computationally more expensive than
the free energy itself.
